---
title: "Warm-up mini-Report: Mosquito Blood Hosts in \n Salt Lake City, Utah"
author: "Gretchen Hellstern"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Abstract

Mosquitoes are a common vector of West Nile Virus. Populations of moquitoes in Salt Lake City, Utah have been showing groupings of WNV. The spread of WNV seems to be aided by many animals around the city. House finches are a common bird species in SLC. The relationship between these two species and the spread of WNV is the focus of our study. We aim to determine if house finches are an amplifying host of WNV. Using a common DNA collection technique, PCR, and sequencing, we were able to obtain a data set of species that had been bitten by mosquitoes, some containing WNV and some not. We gained a large collection of house finch data, which helped confirm our hypothesis that they are amplifying hosts in the spread of WNV. This is supported by our data comparing the amount of each species found at each test location, the detection of WNV in their blood, and the statistical sigificance of the correlation between house finches and WNV.

# Background

West Nile Virus is a common virus in the USA that is spread by the insect vector of the mosquito. Female mosquitoes bite birds and mammals to gain blood meal needed to foster young. In the process of obtaining blood meal, mosquitoes can transfer viruses to the host they've bitten. If that host gets bitten again, that mosquito can than move that virus to another host, increasing the spread of the virus. House finches are among the susceptible hosts. In a recent study, it was found that out of a number of common birds, house finches can harbor the virus in their blood longer than most. This viremia duration barplot (Kumar et al., 2003) illustrates the potential importance of house finches in WNV transmission and as the logical foundation for our hypothesis that house finches serve as amplifying hosts for WNV. This also supports ourprediction that locations with more house finches in our blood host analysis are also the same locations with higher positive tests for WNV in mosquito pools.

```{r viremia, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# Choose some colors
cols <- c(rainbow(30)[c(10:29,1:5)])  # rainbow colors

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

# Study Question and Hypothesis

## Questions  
What role are house finches playing in the spread of WNV?

## Hypothesis 
House finches are acting as important amplifying hosts of WNV in Salt Lake City.

## Prediction
If house finches are acting as important amplifying hosts, we predict that trapping locations where mosquitoes feed on house finches will also have higher rates of confirmed WNV in tested mosquito pools.

# Methods

To begin this experiment, we first collected mosquitoes from multiple traps placed in various locations. We then collected the blood meal from the mosquitoes to use to perform a standard PCR test to amplify the DNA present in the blood meal of each mosquito. We sequenced the DNA samples and used the BLASTn software to identify the species the blood meal came from. We used the species with the highest e-value as our results, as those are the ones the DNA most closely matched.

## First analysis

We compared the number of mosquito blood meals from each host species between the sites with no WNV detected and the sites with WNV detected. The bar plot format helps us to visualize the differences between species, as well as between the two sites.

```{r first-analysis, warning=FALSE}
## import counts_matrix: data.frame with column 'loc_positives' (0/1) and host columns 'host_*'
counts_matrix <- read.csv("./bloodmeal_plusWNV_for_BIOL3070.csv")

## 1) Identify host columns
host_cols <- grep("^host_", names(counts_matrix), value = TRUE)

if (length(host_cols) == 0) {
  stop("No columns matching '^host_' were found in counts_matrix.")
}

## 2) Ensure loc_positives is present and has both levels 0 and 1 where possible
counts_matrix$loc_positives <- factor(counts_matrix$loc_positives, levels = c(0, 1))

## 3) Aggregate host counts by loc_positives
agg <- stats::aggregate(
  counts_matrix[, host_cols, drop = FALSE],
  by = list(loc_positives = counts_matrix$loc_positives),
  FUN = function(x) sum(as.numeric(x), na.rm = TRUE)
)

## make sure both rows exist; if one is missing, add a zero row
need_levels <- setdiff(levels(counts_matrix$loc_positives), as.character(agg$loc_positives))
if (length(need_levels)) {
  zero_row <- as.list(rep(0, length(host_cols)))
  names(zero_row) <- host_cols
  for (lv in need_levels) {
    agg <- rbind(agg, c(lv, zero_row))
  }
  ## restore proper type
  agg$loc_positives <- factor(agg$loc_positives, levels = c("0","1"))
  ## coerce numeric host cols (they may have become character after rbind)
  for (hc in host_cols) agg[[hc]] <- as.numeric(agg[[hc]])
  agg <- agg[order(agg$loc_positives), , drop = FALSE]
}

## 4) Decide species order (overall abundance, descending)
overall <- colSums(agg[, host_cols, drop = FALSE], na.rm = TRUE)
host_order <- names(sort(overall, decreasing = TRUE))
species_labels <- rev(sub("^host_", "", host_order))  # nicer labels

## 5) Build count vectors for each panel in the SAME order
counts0 <- rev(as.numeric(agg[agg$loc_positives == 0, host_order, drop = TRUE]))
counts1 <- rev(as.numeric(agg[agg$loc_positives == 1, host_order, drop = TRUE]))

## 6) Colors: reuse your existing 'cols' if it exists and is long enough; otherwise generate
if (exists("cols") && length(cols) >= length(host_order)) {
  species_colors <- setNames(cols[seq_along(host_order)], species_labels)
} else {
  species_colors <- setNames(rainbow(length(host_order) + 10)[seq_along(host_order)], species_labels)
}

## 7) Shared x-limit for comparability
xmax <- max(c(counts0, counts1), na.rm = TRUE)
xmax <- if (is.finite(xmax)) xmax else 1
xlim_use <- c(0, xmax * 1.08)

## 8) Plot: two horizontal barplots with identical order and colors
op <- par(mfrow = c(1, 2),
          mar = c(4, 12, 3, 2),  # big left margin for species names
          xaxs = "i")           # a bit tighter axis padding

## Panel A: No WNV detected (loc_positives = 0)
barplot(height = counts0,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (-)",
        xlim = xlim_use)

## Panel B: WNV detected (loc_positives = 1)
barplot(height = counts1,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (+)",
        xlim = xlim_use)

par(op)

## Keep the colors mapping for reuse elsewhere
host_species_colors <- species_colors
```

## Second analysis

We then statistically tested whether the presence of house finch blood meal predicts the presence of WNV or not. We are able to gain a p-value, which tells us if our results are statistically significant or not.

```{r}
# second-analysis-or-plot, glm with house finch alone against binary +/_
glm1 <- glm(loc_positives ~ host_House_finch,
            data = counts_matrix,
            family = binomial)
summary(glm1)

#glm with house-finch alone against positivity rate
glm2 <- glm(loc_rate ~ host_House_finch,
            data = counts_matrix)
summary(glm2)
```

# Discussion

Based off of the results we collected, we are able to support our hypothesis that house finches act as important amplifying hosts in the Salt Lake City area.

## Interpretation of 1st analysis

Our first analysis showed that not only is the count of house finches higher at areas with WNV, but the count of house finches at either location is the highest of any species. This shows that they are already favored by mosquitoes to some degree, so they would aide in the spread of WNV accidentally anyways.

## Interpretation of 2nd analysis

Our second analysis shows that the presence of house finch blood meal has a statistically significant correlation to the presence of WNV. We had a p-value of 4.54E-5, which confirms that our results are statiscally significant and are not due to chance.

# Conclusion

Based on the data we collected, our results indicate that house finches play an important role in the spread of WNV in Salt Lake City. This evidence supports our hypothesis that house finches play an amplifying role in the spread of WNV, and that the presence of house finches indicates that WNV is most likely present. This can help us in targeting areas with a dense population of house finches and trying to control the growth of mosquito populations in that area. We can also inform our citizens to be weary of areas containing house finches.

# References    

1. Komar N, Langevin S, Hinten S, Nemeth N, Edwards E, Hettler D, Davis B, Bowen R, Bunning M. Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerg Infect Dis. 2003 Mar;9(3):311-22. https://doi.org/10.3201/eid0903.020628

2. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.